---
title: "St558Project1"
author: "Logan Moy"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(ggplot2)
library(httr)
library(jsonlite)
library(lubridate)
```

# Introduction

SADDASDSSDASDSDSADASD


## Data Processing from PUMS

Function that grabs variables from JSON. Looks at the variables: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION

```{r}
varNames <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION",  "DIVISION")
getVars <- function(url) {
  varInfo <- GET(url)
  parsed <- fromJSON(rawToChar(varInfo$content))
  vars <- parsed$variables
  vars <- vars[names(vars) %in% varNames]
  return (vars)
}
```

Using the function to get variable names from years 2021-2024

```{r}
# Vars from 2021
var2021 <- getVars("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Vars from 2022
var2022 <- getVars("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Vars from 2023
var2023 <- getVars("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Vars from 2024
var2024 <- getVars("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
```

Saving all lists in RDS

```{r}
listVars <- list(var2021, var2022, var2023, var2024)
saveRDS(listVars, "data/variableNames")
```

Making a function to parse API data

```{r}
# Loading the RDS, probably could have put this in a function, but I would have to pass it through the helpers...
varData <- readRDS("data/variableNames")

# Helper function that takes url and gives us the content as a tibble
getUrlInfo <- function(url) {
  dataInfo <- GET(url)
  parsed <- fromJSON(rawToChar(dataInfo$content))
  return (as_tibble(parsed))
}

# Helper for converting Census time codes to midpoint times
parse_time_midpoint <- function(x) {
  x <- as.character(x)
  sapply(x, function(val) {
    # Missing values
    if (is.na(val) || val %in% c("b", "bb", "bbbb", "N.A.", "")) {
      return(NA)
    }
    val <- sprintf("%04d", as.integer(val))
    hour <- as.numeric(substr(val, 1, 2))
    minute <- as.numeric(substr(val, 3, 4))
    hm(sprintf("%02d:%02d", hour, minute))
  })
}

# Helper function for selecting numeric vars in pums
# @param pums tibble we selecting vars from
# @param varNumeric string vector of numeric vars we want
# @throws error when varNumeric does not contain numeric vars
numVarSelect <- function(pums, varNumeric) {
  # Check if varNumeric contains legal variables
  legal <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!(all(varNumeric %in% legal))) {
    stop("varNumeric contains variables that aren't numeric.")
  }
  time_vars <- c("JWAP", "JWDP")
  # Loop that makes varNumeric variables in pums numeric, also does time midpoint
  for (v in varNumeric) {
    if (v %in% time_vars) {
      pums[[v]] <- parse_time_midpoint(pums[[v]])

    } else {
      pums[[v]] <- as.numeric(pums[[v]])
    }
  }
  return (pums)
}

# Helper function for selecting categorical vars in pums
# @param pums tibble we selecting vars from
# @param year we selecting vars for
# @param varCategorical string vector of categorical vars we want
# @throws error when varCategorical does not contain categorical vars
catVarSelect <- function(pums, year, varCategorical) {
  # Check if varCategorical contains legal variables
  legal <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (!(all(varCategorical %in% legal))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  yearVar <- varData[[year - 2020]]
  # Loop that makes varCategorical variables in pums categorical
  for (v in varCategorical) {
    # values and items takes you to factor levels (numbers)
    pums[[v]] <- factor(pums[[v]], levels = names(yearVar[[v]]$values$item), labels = unlist(yearVar[[v]]$values$item))
  }
  return (pums)
}


# Parses the data, uses above helper functions
# @param url string of the url
# @param year int the year, default 2024
# @param varNumeric string a vector of numeric variables to return, default c("AGEP", "PWGTP") 
# @param varCategorical string a vector of categorical variables to return, default c("SEX")
# @returns a tibble with the specified information
# @throws error when year is not 2021-2024
# @throws error when varNumeric does not contain PWGTP or length < 2
# @throws error when varCategorical length == 0
#
getPUMSData <- function(url, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX") ) {

  # Check if the year is valid
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Check if the varNumeric is valid
  if ((!("PWGTP" %in% varNumeric)) | (length(varNumeric) < 2)) {
    stop("Invalid numeric variables given. Vector must be made of strings. Vector must contain PWGTP and have length >= 2.")
  }
  
  # Check if the varCategorical is valid
  if (length(varCategorical)  == 0) {
    stop("Vector must be made of strings. Vector must have length > 0.")
  }
  
  # Getting the tibble
  pums <- getUrlInfo(url)
  
  # Dropping things we dont need
  wanted <- c(varNumeric, varCategorical)
  pums <- pums[, wanted, drop = FALSE]
  
  # Numeric variables to be returned
  pums <- numVarSelect(pums, varNumeric)
  
  # Categorical variables to be returned
  pums <- catVarSelect(pums, year, varCategorical)
  
  class(pums) <- c("census", class(pums))
  return (pums)
}

```


Using the function to get data from years 2021-2024

```{r}
# Data from 2021

# Data from 2022

# Data from 2023

# Data from 2024
```



```{r}
#| echo: false
2 * 2
```

