---
title: "St558Project1"
author: "Logan Moy"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(ggplot2)
library(httr)
library(jsonlite)
library(lubridate)
```

# Introduction

SADDASDSSDASDSDSADASD


## Data Processing from PUMS

```{r}
#| include: false
#| # Added later, wasnt sure we needed a key so never passed it in functions
censusKey <- "aa4f981daae426a93d33caa6ea44623c6ad2d353"
```


### Function that grabs variables from JSON 

Looks at the variables: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION

```{r}
varNames <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION",  "DIVISION")
getVars <- function(url) {
  
  varInfo <- GET(url)
  parsed <- fromJSON(rawToChar(varInfo$content))
  vars <- parsed$variables
  vars <- vars[names(vars) %in% varNames]
  return (vars)
}
```

Using the function to get variable names from years 2021-2024

```{r}
if (!file.exists("data/variableNames")) {
  # Vars from 2021
  var2021 <- getVars("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")
  
  # Vars from 2022
  var2022 <- getVars("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")
  
  # Vars from 2023
  var2023 <- getVars("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")
  
  # Vars from 2024
  var2024 <- getVars("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
  
  #Saving all lists in RDS
  listVars <- list(var2021, var2022, var2023, var2024)
  saveRDS(listVars, "data/variableNames")
}
```

### Making a function to parse API data

```{r}
# Loading the RDS, probably could have put this in a function, but I would have to pass it through the helpers...
varData <- readRDS("data/variableNames")

# Helper function that makes sure we have correct geo variables
# @param int year the year we are interested in
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @returns a list containing geoLvl and geoVal
geoValidator <- function(year, geoLvl, geoVal) {
  
  # Only one geography level allowed
  if (length(geoLvl) != 1) {
    stop("Only one geography level allowed.")
  }
  
  # Check if the geoLvl is valid
  legal <- c("REGION", "DIVISION", "STATE")
  if (!(toupper(geoLvl) %in% legal)) {
    stop("Invalid geoLvl given. Valid options are Region, Division, or State.")
  }

  yearVar <- varData[[year - 2020]]

  # Making sure geoLvl is caps and deals with Year 2021 and 2022 case
  geoLvl <- switch(
    geoLvl,
    REGION   = "REGION",
    DIVISION = "DIVISION",
    STATE    =  "STATE" 
  )

  # Obtain valid values from RDS, names are numbers and its value is the label
  validVals <- names(yearVar[[geoLvl]]$values$item)

  # If we use default geoVal then use the first of the valid values
  if (is.null(geoVal)) {
    geoVal <- validVals[1]
  }

  # All option
  if (toupper(geoVal) == "ALL") {
    return (list(geoLvl = geoLvl, geoVal = NULL))
  }

  # Make sure the geoVal is actually a valid option
  if (!(geoVal %in% validVals)) {
    stop(paste("Invalid value for", geoLvl, ". Must be one of:", paste(validVals, collapse = ", ")))
  }
  return (list(geoLvl = geoLvl, geoVal = geoVal))
}

# Helper function for filling out the URL by pasting info
# Example of URL http://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @param year int year we interestede in
# @param varNumeric vector of strings numeric variables we want
# @param varCategorical vector of strings categorical variables we want
# returns url string which gets used by other functions to retrieve data
createURL <- function(geoLvl, geoVal, year, varNumeric, varCategorical) {
  
  # Variables we want with commas btwn for url
  vars <- paste(c(varNumeric, varCategorical), collapse = ",")
  
  # Overwriting geoVal NULL with * for all
  geoVal <- if (is.null(geoVal)) "*" else geoVal

  # Pasting
  urlString <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    vars,
    "&for=",
    geoLvl,
    ":",
    geoVal,
    "&key=",
    censusKey
  )
  return (urlString)
}


# Helper function that takes url and gives us the content as a tibble
getUrlInfo <- function(urlString) {
  
  dataInfo <- GET(urlString)
  parsed <- fromJSON(rawToChar(dataInfo$content))
  
  # Col names are the first to appear, need to use subsetting to get them
  parsedNames <- parsed[1, ]
  parsed <- parsed[-1, , drop = FALSE]
  colnames(parsed) <- parsedNames
  
  parsed <- as_tibble(parsed)
  
  return (parsed)
}

# Helper for converting Census time codes to midpoint times
parse_time_midpoint <- function(x) {
  
  x <- as.character(x)
  
  sapply(x, function(val) {
    # Missing values
    if (is.na(val) || val %in% c("b", "bb", "bbbb", "N.A.", "")) {
      return(NA)
    }
    
    val <- sprintf("%04d", as.integer(val))
    hour <- as.numeric(substr(val, 1, 2))
    minute <- as.numeric(substr(val, 3, 4))
    hm(sprintf("%02d:%02d", hour, minute))
  })
}

# Helper function for fixing the numerical variables
# @param pums tibble we selecting vars from
# @param varNumeric string vector of numeric vars we want
# @return pums tibble with the correct numerical variables
# @throws error when varNumeric does not contain numeric vars
fixNum <- function(pums, varNumeric) {
  
  time_vars <- c("JWAP", "JWDP")
  
  # Loop that makes varNumeric variables in pums numeric, also does time midpoint
  for (v in varNumeric) {
    if (v %in% time_vars) {
      pums[[v]] <- parse_time_midpoint(pums[[v]])

    } else {
      pums[[v]] <- as.numeric(pums[[v]])
    }
  }
  return (pums)
}

# Helper function for fixing the categorical variables
# @param pums tibble we selecting vars from
# @param year int we selecting vars for
# @param varCategorical string vector of categorical vars we want
# @return pums tibble with the correct categorical variables
# @throws error when varCategorical does not contain categorical vars
fixCate <- function(pums, year, varCategorical) {
  
  yearVar <- varData[[year - 2020]]
  
  # Loop that makes varCategorical variables in pums categorical
  for (v in varCategorical) {
    # values and items takes you to factor levels (numbers)
    pums[[v]] <- factor(pums[[v]], levels = names(yearVar[[v]]$values$item), labels = unlist(yearVar[[v]]$values$item))
  }
  return (pums)
}


# Parses the data, uses above helper functions.
# Gives us a tibble of data from Census API. 
# Subsets the data based on geography, year, and vars
# @param geoLvl string the geography level we want, default is STATE
# @param geoVal string the value of the geography level
# @param year int the year, default 2024
# @param varNumeric string a vector of numeric variables to return, default c("AGEP", "PWGTP") 
# @param varCategorical string a vector of categorical variables to return, default c("SEX")
# @returns a tibble with the specified information
# @throws error when year is not 2021-2024
# @throws error when varNumeric does not contain PWGTP or length < 2
# @throws error when varCategorical length == 0
getYearData <- function(geoLvl = "STATE", geoVal = NULL, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX") ) {

  # Check if the year is valid
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Check if the varNumeric is valid
  if ((!("PWGTP" %in% varNumeric)) | (length(varNumeric) < 2)) {
    stop("Invalid numeric variables given. Vector must be made of strings. Vector must contain PWGTP and have length >= 2.")
  }
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric.")
  }
  
  # Check if the varCategorical is valid
  if (length(varCategorical)  == 0) {
    stop("Vector must be made of strings. Vector must have length > 0.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  # Validates the geo variables
  validGeo <- geoValidator(year, geoLvl, geoVal)
  
  # Creating URL
  urlString <- createURL(validGeo[[1]], validGeo[[2]], year, varNumeric, varCategorical)
  
  # Getting the tibble
  pums <- getUrlInfo(urlString)
  
  # Numeric variables
  pums <- fixNum(pums, varNumeric)
  
  # Categorical variables
  pums <- fixCate(pums, year, varCategorical)
  
  class(pums) <- c("census", class(pums))
  
  return (pums)
}

```

Using the function to get data from year 2021

```{r}
# Data from 2021
example2021 <- getYearData(geoLvl = "DIVISION", year = 2021, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX")) 
```

### Combining Years function

```{r}
# Allows you to get multiple years worth of data from Census API 
# @param year can be a vector of years
# rest comes from getYearData
getCensusData <- function(geoLvl = "STATE", geoVal = NULL, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX")) {

  # Check years
  if (!all(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Better to use some type of apply function, but not covered yet
  pums <- getYearData(geoLvl, geoVal, year[1], varNumeric, varCategorical)
  pums$YEAR <- year[1]
  dataList <- list(pums)
  
  # Starting from 2nd year in vector of years, make sure there is more than 1 year first
  if (length(year) >= 2) {
    for (i in 2:length(year)) {
      pums <- getYearData(geoLvl, geoVal, year[i], varNumeric, varCategorical)
      pums$YEAR <- year[i]
      dataList[[length(dataList) + 1]] <- pums
    }
  }

  # Combine all years
  final <- bind_rows(dataList)
  final$YEAR <- factor(final$YEAR)
  class(final) <- c("census", class(final))

  return (final)
}

```

Example of multiple years function

```{r}
# Data from 2023-4
example20234 <- getCensusData( geoVal = "49", year = c(2023, 2024), varNumeric = c("PWGTP", "GRPIP"), varCategorical = c("HHL")) 
```

## Generic function for summarizing

```{r}
# Helper summary function for our census data (numeric)
# Calculates the mean and sd. Uses PWGTP as a weight vector
# Does not look at time variables
# @param census a census tibble 
# @param varNumeric string of numeric variables we want
# @returns a list with means and sd of numeric 
sumCenNum <- function(census, varNumeric) {

  # Vectors to store means and sds, has same name
  means <- numeric(length(varNumeric))
  sds <- numeric(length(varNumeric))
  names(means) <- varNumeric
  names(sds) <- varNumeric

  # Calculating the summary stats
  for (v in varNumeric) {
    mean <- sum(census[[v]] * census$PWGTP) / sum(census$PWGTP)
    sd <- sqrt(sum(census[[v]]^2 * census$PWGTP) / sum(census$PWGTP) - w_mean^2)

    means[v] <- mean
    sds[v] <- sd
  }

  return (list(means = means,sd = sds))
}

# Helper summary function for our census data (categorical)
# @param census a census tibble 
# @param varCategorical string of numeric variables we want
# @returns a list with counts
sumCenCat <- function(census, varCategorical) {
  
  counts <- list()

  for (v in varCategorical) {
    lvl <- levels(census[[v]])
    countLvl <- numeric(length(lvl))
    names(countLvl) <- lvl
    # Summing for every level of a factor based on weight
    for (l in lvl) {
      countLvl[l] <- sum(census$PWGTP[census[[v]] == l], na.rm = TRUE)
    }
    counts[[v]] <- countLvl
  }
  return (counts)

}

# Summary function for our census data
# @param census a census tibble coming from getCensusData() (as it has year variable)
# @param varNumeric string of numeric variables we want, default is all
# @param varCategorical string of categorical variables we want, default is all
# @returns a list with means and sd of numeric and counts of categorical
# @throws error when vector of variables contains variables not in census or when using
#   weight or time variables
summary.census <- function(census, varNumeric = c("AGEP", "GASP", "GRPIP", "JWMNP"), varCategorical = c("FER", "HHL", "SCH", "SCHL", "SEX", "YEAR")) {
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWMNP")
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric. Or it contains time or weight variables.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX", "YEAR")
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  summaryC <- sumCenNum(census, varNumeric)
  summaryC$count <- sumCenCat(census, varCategorical)
  
  return (summaryC)
}
```

## Generic function for plotting

```{r}
# Plots a boxplot given a numeric and a categorical var
# @param census a census tibble coming from getCensusData() (as it has year variable)
# @param varNumeric numeric variables we want
# @param varCategorical categorical variable we want
# @returns boxplot
# @throws error when we have more than 1 num or cate var
# @throws error when error when vector of variables contains variables not in census or when using
#   weight or time variables
plot.census <- function(census, varNumeric, varCategorical) {
  
  # Check if only 1 var is used
  if (length(varNumeric) != 1 | length(varCategorical) != 1) {
    stop("Only 1 numeric and categorical variable can be used.");
  }
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWMNP")
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric. Or it contains time or weight variables.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX", "YEAR")
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  
  
  # Boxplot
  ggplot(census, aes(x = .data[[varCategorical]], y = .data[[varNumeric]], weight = PWGTP)) + geom_boxplot() + 
    labs(x = varCategorical, y = varNumeric, title = paste(varNumeric, "by", varCategorical)) + 
    theme_minimal()
  
}
```

# Example: Investigating the _____________________
```{r}
ggplot(example20234 ) + geom_boxplot(aes(x = HHL, y = .data[['GRPIP']], weight = PWGTP)) + 
  theme_minimal()
```


