---
title: "St558Project1"
author: "Logan Moy"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(ggplot2)
library(httr)
library(jsonlite)
library(lubridate)
library(quantreg)
```

# Introduction

For this assignment we will investigate the data within the PUMS Census API. PUMS includes information about America's economy and population. We'll examine data from years 2021-2024 and look at the variables AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST/STATE , REGION, and DIVISION.

First we need some functions to actually grab the data from that API. Since the data is in a JSON format we'll need to use functions in httr and jsonlite package to obtain the data. 

Since the API gives us information about the variables in it, we'll read those in as well. This helps us get information like factor names for categorical variables.


# Data Processing from PUMS

```{r}
#| include: false
#| # Added later, wasnt sure we needed a key so never passed it in functions
censusKey <- "aa4f981daae426a93d33caa6ea44623c6ad2d353"
```


## Function that grabs variables from JSON 

```{r}
varNames <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION",  "DIVISION")
getVars <- function(url) {
  
  varInfo <- GET(url)
  parsed <- fromJSON(rawToChar(varInfo$content))
  vars <- parsed$variables
  vars <- vars[names(vars) %in% varNames]
  return (vars)
}
```

Using the function to get variable names from years 2021-2024

```{r}
if (!file.exists("data/variableNames")) {
  # Vars from 2021
  var2021 <- getVars("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")
  
  # Vars from 2022
  var2022 <- getVars("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")
  
  # Vars from 2023
  var2023 <- getVars("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")
  
  # Vars from 2024
  var2024 <- getVars("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
  
  #Saving all lists in RDS
  listVars <- list(var2021, var2022, var2023, var2024)
  saveRDS(listVars, "data/variableNames")
}
```

## Making a function to parse API data

Here we'll start reading in the data. Process is broken up into multiple helper functions.

We read in variable name data we stored earlier.

We make sure the geography levels and values are legal.

We have a function that creates url information that makes an url that lets us request the year, geography, and variables we want.

This url is then used by another function which turns the data from that url into a tibble.

Then we make sure the tibble's columns are in the correct format.

All of this gets wrapped up in a wrapper function that gives us that year's information.

```{r}
# Loading the RDS, probably could have put this in a function, but I would have to pass it through the helpers...
varData <- readRDS("data/variableNames")

# Helper function that makes sure we have correct geo variables
# @param int year the year we are interested in
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @returns a list containing geoLvl and geoVal
geoValidator <- function(year, geoLvl, geoVal) {
  
  # Only one geography level allowed
  if (length(geoLvl) != 1) {
    stop("Only one geography level allowed.")
  }
  
  # Check if the geoLvl is valid
  legal <- c("REGION", "DIVISION", "STATE")
  if (!(toupper(geoLvl) %in% legal)) {
    stop("Invalid geoLvl given. Valid options are Region, Division, or State.")
  }

  yearVar <- varData[[year - 2020]]

  # Making sure geoLvl is caps and deals with Year 2021 and 2022 case
  geoLvl <- switch(
    geoLvl,
    REGION   = "REGION",
    DIVISION = "DIVISION",
    STATE    =  if_else(year < 2023, "ST", "STATE") 
  )

  # Obtain valid values from RDS, names are numbers and its value is the label
  validVals <- names(yearVar[[geoLvl]]$values$item)

  # If we use default geoVal then use the first of the valid values
  if (is.null(geoVal)) {
    geoVal <- validVals[1]
  }

  # All option
  if (toupper(geoVal) == "ALL") {
    return (list(geoLvl = geoLvl, geoVal = NULL))
  }

  # Make sure the geoVal is actually a valid option
  if (!(geoVal %in% validVals)) {
    stop(paste("Invalid value for", geoLvl, ". Must be one of:", paste(validVals, collapse = ", ")))
  }
  return (list(geoLvl = geoLvl, geoVal = geoVal))
}

# Helper function for filling out the URL by pasting info
# Example of URL http://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @param year int year we interestede in
# @param varNumeric vector of strings numeric variables we want
# @param varCategorical vector of strings categorical variables we want
# returns url string which gets used by other functions to retrieve data
createURL <- function(geoLvl, geoVal, year, varNumeric, varCategorical) {
  
  # Variables we want with commas btwn for url
  vars <- paste(c(varNumeric, varCategorical), collapse = ",")
  
  # Overwriting geoVal NULL with * for all
  geoVal <- if_else(is.null(geoVal), "*", geoVal)
  
  # State needs to be state not st
  if (toupper(geoLvl) == "ST") {
    geoLvl <- "STATE"
  }

  # Pasting
  urlString <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    vars,
    "&for=",
    geoLvl,
    ":",
    geoVal,
    "&key=",
    censusKey
  )
  return (urlString)
}


# Helper function that takes url and gives us the content as a tibble
getUrlInfo <- function(urlString) {
  
  dataInfo <- GET(urlString)
  parsed <- fromJSON(rawToChar(dataInfo$content))
  
  # Col names are the first to appear, need to use subsetting to get them
  parsedNames <- parsed[1, ]
  parsed <- parsed[-1, , drop = FALSE]
  colnames(parsed) <- parsedNames
  
  parsed <- as_tibble(parsed)
  
  return (parsed)
}

# Helper for converting Census time codes to midpoint times
parse_time_midpoint <- function(x) {
  
  x <- as.character(x)
  
  sapply(x, function(val) {
    # Missing values
    if (is.na(val) || val %in% c("b", "bb", "bbbb", "N.A.", "")) {
      return(NA)
    }
    
    val <- sprintf("%04d", as.integer(val))
    hour <- as.numeric(substr(val, 1, 2))
    minute <- as.numeric(substr(val, 3, 4))
    hm(sprintf("%02d:%02d", hour, minute))
  })
}

# Helper function for fixing the numerical variables
# @param pums tibble we selecting vars from
# @param varNumeric string vector of numeric vars we want
# @return pums tibble with the correct numerical variables
# @throws error when varNumeric does not contain numeric vars
fixNum <- function(pums, varNumeric) {
  
  time_vars <- c("JWAP", "JWDP")
  
  # Loop that makes varNumeric variables in pums numeric, also does time midpoint
  for (v in varNumeric) {
    if (v %in% time_vars) {
      pums[[v]] <- parse_time_midpoint(pums[[v]])

    } else {
      pums[[v]] <- as.numeric(pums[[v]])
    }
  }
  return (pums)
}

# Helper function for fixing the categorical variables
# @param pums tibble we selecting vars from
# @param year int we selecting vars for
# @param varCategorical string vector of categorical vars we want
# @return pums tibble with the correct categorical variables
# @throws error when varCategorical does not contain categorical vars
fixCate <- function(pums, year, varCategorical) {
  
  yearVar <- varData[[year - 2020]]
  
  # Loop that makes varCategorical variables in pums categorical
  for (v in varCategorical) {
    # values and items takes you to factor levels (numbers)
    pums[[v]] <- factor(pums[[v]], levels = names(yearVar[[v]]$values$item), labels = unlist(yearVar[[v]]$values$item))
  }
  return (pums)
}


# Parses the data, uses above helper functions.
# Gives us a tibble of data from Census API. 
# Subsets the data based on geography, year, and vars
# @param geoLvl string the geography level we want, default is STATE
# @param geoVal string the value of the geography level
# @param year int the year, default 2024
# @param varNumeric string a vector of numeric variables to return, default c("AGEP", "PWGTP") 
# @param varCategorical string a vector of categorical variables to return, default c("SEX")
# @returns a tibble with the specified information
# @throws error when year is not 2021-2024
# @throws error when varNumeric does not contain PWGTP or length < 2
# @throws error when varCategorical length == 0
getYearData <- function(geoLvl = "STATE", geoVal = NULL, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX") ) {

  # Check if the year is valid
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Check if the varNumeric is valid
  if ((!("PWGTP" %in% varNumeric)) | (length(varNumeric) < 2)) {
    stop("Invalid numeric variables given. Vector must be made of strings. Vector must contain PWGTP and have length >= 2.")
  }
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric.")
  }
  
  # Check if the varCategorical is valid
  if (length(varCategorical)  == 0) {
    stop("Vector must be made of strings. Vector must have length > 0.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  # Validates the geo variables
  validGeo <- geoValidator(year, geoLvl, geoVal)
  
  # Creating URL
  urlString <- createURL(validGeo[[1]], validGeo[[2]], year, varNumeric, varCategorical)
  
  # Getting the tibble
  pums <- getUrlInfo(urlString)
  
  # Numeric variables
  pums <- fixNum(pums, varNumeric)
  
  # Categorical variables
  pums <- fixCate(pums, year, varCategorical)
  
  class(pums) <- c("census", class(pums))
  
  return (pums)
}

```

### Examples of the getYearData function

Using the function to get data from year 2021

```{r}
example2021 <- getYearData(geoLvl = "DIVISION", year = 2021, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX")) 
```

Using the function to get data from year 2024

```{r}
example2024 <- getYearData( varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = c("SCH", "SCHL", "SEX")) 
```

Proof of Error handling

Handling of case of using incorrect var categorical 

```{r}
tryCatch({
  getYearData(varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = c("SCH", "SCHL", "ASD"))
  },
  error = function(e) {
    geterrmessage() == "varCategorical contains variables that aren't categorical."
  }
) 

tryCatch({
  getYearData(varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = NULL)
  },
  error = function(e) {
    geterrmessage() == "Vector must be made of strings. Vector must have length > 0."
  }
) 

```

Bad year and not using PWGTP

```{r}
tryCatch({
  getYearData(year = 2222, varNumeric = c("GASP", "GRPIP", "JWAP"), varCategorical = c("SCH", "SCHL", "ASD"))
  },
  error = function(e) {
    geterrmessage() == "Invalid year given. Year must be numeric and between 2021-2024 (inclusive)."
  }
) 

tryCatch({
  getYearData(varNumeric = c("GASP", "GRPIP", "JWAP"), varCategorical = NULL)
  },
  error = function(e) {
    geterrmessage() == "Invalid numeric variables given. Vector must be made of strings. Vector must contain PWGTP and have length >= 2."
  }
)
```

Geo lvl error

```{r}
tryCatch({
  getYearData(geoLvl = "asd", varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = "SCHL")
  },
  error = function(e) {
    geterrmessage() == "Invalid geoLvl given. Valid options are Region, Division, or State."
  }
) 

tryCatch({
  getYearData(geoLvl = c("STATE", "REGION"), varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = "SCHL")
  },
  error = function(e) {
    geterrmessage() == "Only one geography level allowed."
  }
) 
```

## Combining Years function

Here is the final form of the function we use to get data from the API. This function uses the previous getYearData and a loop to get data from multiple years. This allows the user to compare two+ years (although this function can still take 1 year).

```{r}
# Allows you to get multiple years worth of data from Census API 
# @param year can be a vector of years
# rest comes from getYearData
getCensusData <- function(geoLvl = "STATE", geoVal = NULL, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX")) {

  # Check years
  if (!all(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Better to use some type of apply function, but not covered yet
  pums <- getYearData(geoLvl, geoVal, year[1], varNumeric, varCategorical)
  pums$YEAR <- year[1]
  dataList <- list(pums)
  
  # Starting from 2nd year in vector of years, make sure there is more than 1 year first
  if (length(year) >= 2) {
    for (i in 2:length(year)) {
      pums <- getYearData(geoLvl, geoVal, year[i], varNumeric, varCategorical)
      pums$YEAR <- year[i]
      dataList[[length(dataList) + 1]] <- pums
    }
  }

  # Combine all years
  final <- bind_rows(dataList)
  final$YEAR <- factor(final$YEAR)
  class(final) <- c("census", class(final))

  return (final)
}

```

### Examples of getCensusData function

```{r}
# Data from 2023-4
example20234 <- getCensusData( geoVal = "49", year = c(2023, 2024), varNumeric = c("PWGTP", "GRPIP"), varCategorical = c("HHL")) 
```


```{r}
# Data from 
example20212024 <- getCensusData(geoLvl = "REGION", geoVal = "2", year = c(2021, 2024), varNumeric = c("PWGTP", "GRPIP", "GASP", "AGEP", "JWAP"), varCategorical = c("HHL", "FER", "SEX")) 
```

Bad years

```{r}
tryCatch({
  getCensusData(year = c(2022, 1111), geoLvl = "REGION", varNumeric = c("GASP", "GRPIP", "JWAP", "PWGTP"), varCategorical = "SCHL")
  },
  error = function(e) {
    geterrmessage() == "Invalid year given. Year must be numeric and between 2021-2024 (inclusive)."
  }
) 
```


# Generic functions

Now that we have some data, we can do some exploratory data analysis on it. We want to see what the data looks like through summaries and plots. We want to look at numeric data's mean and sd. For categorical we'll just look at their counts. Since there is a weight variable, that will be taken into account in the following functions.


## Function for summarizing

```{r}
# Helper summary function for our census data (numeric)
# Calculates the mean and sd. Uses PWGTP as a weight vector
# Does not look at time variables
# @param census a census tibble 
# @param varNumeric string of numeric variables we want
# @returns a list with means and sd of numeric 
sumCenNum <- function(census, varNumeric) {

  # Vectors to store means and sds, has same name
  means <- numeric(length(varNumeric))
  sds <- numeric(length(varNumeric))
  names(means) <- varNumeric
  names(sds) <- varNumeric

  # Calculating the summary stats
  for (v in varNumeric) {
    mean <- sum(census[[v]] * census$PWGTP) / sum(census$PWGTP)
    sd <- sqrt(sum(census[[v]]^2 * census$PWGTP) / sum(census$PWGTP) - mean^2)

    means[v] <- mean
    sds[v] <- sd
  }

  return (list(means = means,sd = sds))
}

# Helper summary function for our census data (categorical)
# @param census a census tibble 
# @param varCategorical string of numeric variables we want
# @returns a list with counts
sumCenCat <- function(census, varCategorical) {
  
  counts <- list()

  for (v in varCategorical) {
    lvl <- levels(census[[v]])
    countLvl <- numeric(length(lvl))
    names(countLvl) <- lvl
    # Summing for every level of a factor based on weight
    for (l in lvl) {
      countLvl[l] <- sum(census$PWGTP[census[[v]] == l], na.rm = TRUE)
    }
    counts[[v]] <- countLvl
  }
  return (counts)

}

# Summary function for our census data
# @param census a census tibble coming from getCensusData() (as it has year variable)
# @param varNumeric string of numeric variables we want, default is all
# @param varCategorical string of categorical variables we want, default is all
# @returns a list with means and sd of numeric and counts of categorical
# @throws error when vector of variables contains variables not in census or when using
#   weight or time variables
summary.census <- function(census, varNumeric = NULL, varCategorical = NULL) {
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWMNP")
  
  if (is.null(varNumeric)) {
    varNumeric <- intersect(legalN, names(census))
  }
  
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric. Or it contains time or weight variables.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX", "YEAR")
  
  if (is.null(varCategorical)) {
    varCategorical <- intersect(legalC, names(census))
  }
  
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  summaryC <- sumCenNum(census, varNumeric)
  summaryC$count <- sumCenCat(census, varCategorical)
  
  return (summaryC)
}
```

### Examples of summary

```{r}
summary(example20234)
```

```{r}
summary(example20212024, "AGEP", "FER")
```
Some errors

```{r}
tryCatch({
  summary(example20212024, "AGEP", "SSS")
  },
  error = function(e) {
    geterrmessage() == "varCategorical contains variables that aren't categorical."
  }
) 
```



## Function for plotting

```{r}
# Plots a boxplot given a numeric and a categorical var
# @param census a census tibble coming from getCensusData() (as it has year variable)
# @param varNumeric numeric variables we want
# @param varCategorical categorical variable we want
# @returns boxplot
# @throws error when we have more than 1 num or cate var
# @throws error when error when vector of variables contains variables not in census or when using
#   weight or time variables
plot.census <- function(census, varNumeric, varCategorical) {
  
  # Check if only 1 var is used
  if (length(varNumeric) != 1 | length(varCategorical) != 1) {
    stop("Only 1 numeric and categorical variable can be used.")
  }
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWMNP")
  if (!(all(varNumeric %in% legalN))) {
    stop("varNumeric contains variables that aren't numeric. Or it contains time or weight variables.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX", "YEAR")
  if (!(all(varCategorical %in% legalC))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  
  
  # Boxplot
  ggplot(census, aes(x = .data[[varCategorical]], y = .data[[varNumeric]], weight = PWGTP)) + geom_boxplot() + 
    labs(x = varCategorical, y = varNumeric, title = paste(varNumeric, "by", varCategorical)) + 
    theme_minimal()
  
}
```

### Example of the plot function

```{r}
plot(example20234, "GRPIP", "HHL")
```

Plot errors

```{r}
tryCatch({
  plot(example20212024, "AGEP", c("FER", "YEAR"))
  },
  error = function(e) {
    geterrmessage() == "Only 1 numeric and categorical variable can be used."
  }
)
```


# Example of Process

Let's say I was interested in the relationship of how much a Kentucky resident pays for gas by their fertility (). People who have given birth recently dont drive as much as does who do.

First I'll give my getCensusData function the geoVal of 21 which is Kentucky (I dont need to give it a State as State is default). Then a vector of years and the numeric variables I want (gas prices and the weight vector) and the categorical variable I am interested in (fertility). I added some extra variables so we can request by name the ones we want later.

```{r}
proex <- getCensusData(geoVal = "21", year = c(2021, 2022, 2023, 2024), varNumeric = c("GASP", "PWGTP", "AGEP", "JWAP"), varCategorical = c("SEX", "FER"))

```

After that I can now look at a summary of the data given back by the function with just a summary() as I made a generic one earlier.

```{r}
summary(proex, varNumeric = "GASP", varCategorical = "FER")
```

Summary looks fine. Gas expenses has a large sd compared to its mean. As for fertility, most people seem to make up the N/A which makes sense as only a portion of females can give birth.

Lets compare their distributions below.


```{r}
plot(proex, "GASP", "FER")
```

Maybe giving birth/being pregnant could have an effect on how much you spend on gas a month. Not being pregnant, N/A and No, seem to have a similar distribution, while Yes could slightly differ graphically. What is strange is that all three categories have the same outliers.

Overall it looks to hard to tell is the distribution differs between fertility status.

