---
title: "St558Project1"
author: "Logan Moy"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(ggplot2)
library(httr)
library(jsonlite)
library(lubridate)
```

# Introduction

SADDASDSSDASDSDSADASD


## Data Processing from PUMS

```{r}
#| include: false
censusKey <- "aa4f981daae426a93d33caa6ea44623c6ad2d353"
```


### Function that grabs variables from JSON 

Looks at the variables: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION

```{r}
varNames <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION",  "DIVISION")
getVars <- function(url) {
  
  varInfo <- GET(url)
  parsed <- fromJSON(rawToChar(varInfo$content))
  vars <- parsed$variables
  vars <- vars[names(vars) %in% varNames]
  return (vars)
}
```

Using the function to get variable names from years 2021-2024

```{r}
# Vars from 2021
var2021 <- getVars("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Vars from 2022
var2022 <- getVars("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Vars from 2023
var2023 <- getVars("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Vars from 2024
var2024 <- getVars("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
```

Saving all lists in RDS

```{r}
listVars <- list(var2021, var2022, var2023, var2024)
saveRDS(listVars, "data/variableNames")
```

### Making a function to parse API data

```{r}
# Loading the RDS, probably could have put this in a function, but I would have to pass it through the helpers...
varData <- readRDS("data/variableNames")

# Helper function that makes sure we have correct geo variables
# @param int year the year we are interested in
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @returns a list containing geoLvl and geoVal
geoValidator <- function(year, geoLvl, geoVal) {
  
  # Only one geography level allowed
  if (length(geography) != 1) {
    stop("Only one geography level allowed.")
  }
  
  # Check if the geoLvl is valid
  legal <- c("REGION", "DIVISION", "STATE")
  if (!(toupper(geoLvl) %in% legal)) {
    stop("Invalid geoLvl given. Valid options are Region, Division, or State.")
  }

  yearVar <- varData[[year - 2020]]

  # Making sure geoLvl is caps and deals with Year 2021 and 2022 case
  geoLvl <- switch(
    geoLvl,
    Region   = "REGION",
    Division = "DIVISION",
    State    = if (year <= 2022) "ST" else "STATE"
  )

  # Obtain valid values from RDS, names are numbers and its value is the label
  validVals <- names(yearVar[[geoLvl]]$values$item)

  # If we use default geoVal then use the first of the valid values
  if (is.null(geoVal)) {
    geoVal <- validVals[1]
  }

  # All option
  if (toupper(geoVal) == "ALL") {
    return (list(geoLvl = geoLvl, geoVal = NULL))
  }

  # Make sure the geoVal is actually a valid option
  if (!(geoVal %in% validVals)) {
    stop(paste("Invalid value for", geography, ". Must be one of:", paste(validVals, collapse = ", ")))
  }
  return (list(geoLvl = geoLvl, geoVal = geoVal))
}

# Helper function for filling out the URL by pasting info
# Example of URL http://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @param year int year we interestede in
# @param varNumeric vector of strings numeric variables we want
# @param varCategorical vector of strings categorical variables we want
# returns url string which gets used by other functions to retrieve data
createURL <- function(geoLvl, geoVal, year, varNumeric, varCategorical) {
  
  # Starting URL
  url <- "https://api.census.gov/data/"
  
  # Adding year
  url <- paste0(url, year, "/acs/acs1/pums?get=")
  
  # Variables to use and adding geo lvl to url
  vars <- paste(c(varNumeric, varCategorical), collapse = ",")    
  url <- paste0(url, vars, "&for=", geoLvl, ":")
  
  # Adding geoVal
  if (is.null(geoVal)) {
    url <- paste0(url, "*")
  } else {
    url <- paste0(url, geoVal)
  }
  
  # Adding the key
  url <- paste0(url, "&key=", censusKey)
  return (url)
}

# Helper function that takes url and gives us the content as a tibble
getUrlInfo <- function(url) {
  
  dataInfo <- GET(url)
  parsed <- fromJSON(rawToChar(dataInfo$content))
  
  return (as_tibble(parsed))
}

# Helper for converting Census time codes to midpoint times
parse_time_midpoint <- function(x) {
  
  x <- as.character(x)
  
  sapply(x, function(val) {
    # Missing values
    if (is.na(val) || val %in% c("b", "bb", "bbbb", "N.A.", "")) {
      return(NA)
    }
    
    val <- sprintf("%04d", as.integer(val))
    hour <- as.numeric(substr(val, 1, 2))
    minute <- as.numeric(substr(val, 3, 4))
    hm(sprintf("%02d:%02d", hour, minute))
  })
}

# Helper function for fixing the numerical variables
# @param pums tibble we selecting vars from
# @param varNumeric string vector of numeric vars we want
# @return pums tibble with the correct numerical variables
# @throws error when varNumeric does not contain numeric vars
fixNum <- function(pums, varNumeric) {
  
  time_vars <- c("JWAP", "JWDP")
  
  # Loop that makes varNumeric variables in pums numeric, also does time midpoint
  for (v in varNumeric) {
    if (v %in% time_vars) {
      pums[[v]] <- parse_time_midpoint(pums[[v]])

    } else {
      pums[[v]] <- as.numeric(pums[[v]])
    }
  }
  return (pums)
}

# Helper function for fixing the categorical variables
# @param pums tibble we selecting vars from
# @param year int we selecting vars for
# @param varCategorical string vector of categorical vars we want
# @return pums tibble with the correct categorical variables
# @throws error when varCategorical does not contain categorical vars
fixCate <- function(pums, year, varCategorical) {
  
  yearVar <- varData[[year - 2020]]
  
  # Loop that makes varCategorical variables in pums categorical
  for (v in varCategorical) {
    # values and items takes you to factor levels (numbers)
    pums[[v]] <- factor(pums[[v]], levels = names(yearVar[[v]]$values$item), labels = unlist(yearVar[[v]]$values$item))
  }
  return (pums)
}


# Parses the data, uses above helper functions.
# Gives us a tibble of data from Census API. 
# Subsets the data based on geography, year, and vars
# @param geoLvl string the geography level we want
# @param geoVal string the value of the geography level
# @param year int the year, default 2024
# @param varNumeric string a vector of numeric variables to return, default c("AGEP", "PWGTP") 
# @param varCategorical string a vector of categorical variables to return, default c("SEX")
# @returns a tibble with the specified information
# @throws error when year is not 2021-2024
# @throws error when varNumeric does not contain PWGTP or length < 2
# @throws error when varCategorical length == 0
getPUMSData <- function(geoLvl = "State", geoVal = NULL, year = 2024, varNumeric = c("AGEP", "PWGTP"), varCategorical = c("SEX") ) {

  # Check if the year is valid
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Invalid year given. Year must be numeric and between 2021-2024 (inclusive).")
  }
  
  # Check if the varNumeric is valid
  if ((!("PWGTP" %in% varNumeric)) | (length(varNumeric) < 2)) {
    stop("Invalid numeric variables given. Vector must be made of strings. Vector must contain PWGTP and have length >= 2.")
  }
  
  # Check if varNumeric contains legal variables
  legalN <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!(all(varNumeric %in% legal))) {
    stop("varNumeric contains variables that aren't numeric.")
  }
  
  # Check if the varCategorical is valid
  if (length(varCategorical)  == 0) {
    stop("Vector must be made of strings. Vector must have length > 0.")
  }
  
  # Check if varCategorical contains legal variables
  legalC <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (!(all(varCategorical %in% legal))) {
    stop("varCategorical contains variables that aren't categorical.")
  }
  
  # Validates the geo variables
  validGeo <- geoValidator(year, geoLvl, geoVal)
  
  # Creating URL
  urlString <- createURL(validGeo[[1]], validGeo[[2]], year, varNumeric, varCategorical)
  
  # Getting the tibble
  pums <- getUrlInfo(url)
  
  # Numeric variables
  pums <- fixNum(pums, varNumeric)
  
  # Categorical variables
  pums <- fixCate(pums, year, varCategorical)
  
  class(pums) <- c("census", class(pums))
  
  return (pums)
}

```


Using the function to get data from years 2021-2024

```{r}
# Data from 2021

# Data from 2022

# Data from 2023

# Data from 2024
```

### Combining Years function

```{r}
#| echo: false
2 * 2
```


## Generic function for summarizing

```{r}

```



# Example: Investigating the _____________________


