---
title: "St558Project1"
author: "Logan Moy"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(ggplot2)
library(httr)
library(jsonlite)
```

# Introduction

SADDASDSSDASDSDSADASD


## Data Processing from PUMS

Function that grabs variables from JSON. Looks at the variables: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION

```{r}
varNames <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION",  "DIVISION")
getVars <- function(url) {
  varInfo <- GET(url)
  parsed <- fromJSON(rawToChar(varInfo$content))
  vars <- parsed$variables
  vars <- vars[names(vars) %in% varNames]
  return (vars)
}
```

Using the function to get variable names from years 2021-2024

```{r}
# Vars from 2021
var2021 <- getVars("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Vars from 2022
var2022 <- getVars("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Vars from 2023
var2023 <- getVars("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Vars from 2024
var2024 <- getVars("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
```

Saving all lists in RDS

```{r}
variableNames <- list(var2021, var2022, var2023, var2024)
saveRDS(variableNames, "data/variableNames")
```

Making a function to parse API data

```{r}

```


Using the function to get data from years 2021-2024

```{r}
# Data from 2021

# Data from 2022

# Data from 2023

# Data from 2024
```



```{r}
#| echo: false
2 * 2
```

